### 2014-10-06 paper/chemokine_main_revised.tex
middle of Steph's edits

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{ulem}

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a small fraction of the total lung.  T cell navigation towards infected foci is undirected, but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.  We use a spatially explicit agent-based model (ABM) to examine the efficiency of the recruitment of activated CD8 T cells by local chemokine signaling.

To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.  These data were used to estimate chemokine production rates. The estimates were combined in the spatial model to test viral variation of T cell recruitment in the lungs.

The spatial nature of the ABM model reveals unique challenges to T cell recruitment not appreciated in standard differential equation models.  Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.  This spatially explicit model describes efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.  A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in finding infected cells in large foci.  Finally, a sensitivity analysis highlights the viral parameters as the most significant in influencing the result of the infection.  Thus, we conclude that viral dynamics, rather than chemokine dynamics, dominate the efficiency of T cell recruitment in determining whether the influenza infection is controlled.

@@

%The adaptive immune response to influenza A infection in the lung requires a complex combination of different defensive mechanisms.  CD8 T cells must navigate from the lymph node through the vascular system to the site of infection.  
%
%
%The high replication rate of 2009 pandemic H1N1 influenza presents unique challenges to the immune system when compared to other strains of influenza.  Studies have attributed the high mortality rate of hosts infected with pH1N1 to the virus' high replication rate.
%
%We suggest pH1N1 virulence may only tell part of the story.  Static analysis of viral production rates and plaque sizes ignore the spatial structure of the infection.  We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain appropriate numbers of activated CD8 T cells.  
%
%There are many mathematical models of the adaptive response \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.  Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail \cite{Beauchemin2006a, Bauer2009}.   Here we examine spatial aspects of the the CD8 T cell diaspora and recruitment to infected lung.  Activated CD8 T cells searching for and homing into infected tissue do so in a complex environment.  We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.  In the model, CD8 T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.  
%
%%Therefore, we constructed a spatial agent-based model to examine how T cells are recruited to the site of infection, and once there, how they navigate over the plaque to contain the growing infection.  Because our model strives to answer the specific questions regarding T cell recruitment and movement in the lung, we ignore other important aspects of the immune response such as...
%
%% Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, and regulatory interactions with other cell phenotypes \cite{McGill2009}.  The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.  
%
%Previous work has shown that T cell search is likely aided by chemokine secretion by infected cells \cite{Banerjee2011}, but details of this process are unknown.  To find secretion rates of chemokines, we infected human epithelial cells \textit{in virto} with one of three strains of influenza: avian aH5N1, seasonal sH1N1, and pandemic 2009 pH1N1.  Wells were measured  for specific chemokines at specific time points over a period of two days.  Two chemokines: IP-10 (CXCL10) and RANTES (CCL5) displayed elevated concentrations over the course of the infections.  We extended a previous ODE model \cite{Mitchell2011} and fit it to the chemokine data to deduce per cell chemokine production rates.  We then used the resulting values to inform our spatial model of T cell recruitment.
%
%Our results reveal unique spatial challenges to T cell search that arise from chemokine recruitment of T cells to the site of infection.  Molecular diffusion of chemokines creates the chemical gradients that allow T cells to find the infection.  Conversely, the coarse nature of the diffusion process creates pockets of high chemokine concentration.  These pockets, or local maxima, have the unfortunate effect of recruiting and trapping T cells that navigate within their basin of attraction.  This effect hinders the ability of T cells to properly cover the entire infection.  We find that viruses with very high replication rates, such as pandemic H1N1, can create infections that spread across tissue faster than the chemokine signals can diffuse.  In these instances, the chemical signal that is so crucial to T cell recruitment will actually hinder the immune system's search process.
%
%These results show interventions that increase chemokine concentrations and T cell counts may not help clear an infection.  Furthermore, clearing the pH1N1 infection may be limited by the rate at which T cells can kill infected epithelial cells.  This suggests that early interventions to infection, such as vaccination, may be the only way to avoid runaway infection.

The adaptive immune response induced during acute influenza A infection is an integrated complex combination of defense mechanisms.  Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control the infection and immunopathology.   The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.   Complete resolution of influenza pneumonia depends on antigen-specific CD8+ T effector cells \cite{Cerwenka1999, Kim2011}.  The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.   On approximately day 5 after infection, activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.

CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in infected lung.  Search problems in the lymph node have been simulated using live cell imaging data to provides reliable parameters of cell movement \cite{Vroomans2012}.  The T precursor cell must encounter on average hundreds of irrelevant cells before it contacts the loaded DC, but the distance traveled is relatively short, and the direction is likely guided by chemokine signals.

The second search, i.e. the process of recruiting activated CD8 effector cells to infected sites in the lung,  is different and not as well understood.  In contrast to the dense lymph node volume, lung infection dynamics occur on the surface of epithelial cells, the size of the tissue layer being comparable to the size of a tennis court.  Initial localization and extravasation may depend on a number of inflammatory signals but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.   Moreover, the T cells leaving the lymph node are released into the very large systemic vascular `space' with no initial guiding signals.  Effector cells must localize to sites of viral replication \cite{Cerwenka1999}, but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels. Chemotactic proteins play a major role in recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005} where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005} and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.

The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.   For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediates T cell recruitment to infected brain and subsequent immunopathology \cite{Christensen2004, Christensen2006}, while the CCR5 mediates the opposite effect \cite{DeLemos2005}.  In contrast, in the West Nile virus model \cite{Klein2005} and the dengue model \cite{Hsieh2006}, deficiency of IP-10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.   In the herpesvirus model, IP-10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \cite{Wuest2008, Thapa2008}.  In the parainfluenza virus model, CXCR3 receptor is critical in CD4+ T cell migration to the lung \cite{Kohlmeier2009}.  In the influenza A model, initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \cite{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.   The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency, but viral clearance was not altered \cite{Fadel2008}.  The CXCR3 receptor also mediates the balance between effector versus memory cell differentiation among recruited CD8 T cells in the lung \cite{Kohlmeier2011}.

Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.   Following arrival at the infected site, CD8 T cells secrete cytokines when mononuclear cells are present but in their absence kill infected epithelial cells only by contact cytolysis \cite{Hufford2011}.   Finally the exit of T cells from tissue is tightly regulated to keep effector cells from leaving or dying prematurely \cite{Jennrich2012, Richter2007} but effective life-span is unknown.

There are many mathematical models of the entire adaptive response \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.  Mathematical and computational models can examine each event in detail and narrow the range of possible behaviors.  Here we examine the CD8 T cell diaspora and recruitment to infected lung in detail.  Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail \cite{Beauchemin2006a, Bauer2009}.   Activated CD8 T cells searching for and homing into infected tissue do so in a spatially complex environment.  We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.  In the model, CD8 T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.

We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.  Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, and regulatory interactions with other cell phenotypes \cite{McGill2009}.  The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.

Previously we developed an agent-based model of the lymphocyte diaspora focusing on the production and diffusion of a generic inflammatory signal without experimental data to calibrate the model \cite{Banerjee2011}.  Here we refine that model and incorporate chemotactic protein data from in vitro cultures of human bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} to focus on two features of searching in the large lung space.  First, functional thresholds of the chemotactic proteins were surprisingly low in our model.  Second there appears to be a threshold size of the infected focus above which the T cell search becomes inefficient.

@@

@@

Epithelial cell culture and supernatant collection was performed as described in \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.  TNF activity as not incorporated in the model.  Data for other chemokines and cytokines is not shown.

\subsection*{Computational Modeling}

We implemented our spatial model using CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.  A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}) and simulations measured efficiency of infection clearance for three different viruses. The lung was represented as a two-dimensional sheet of healthy epithelial cells (S1.2).  Because we assume activated T cells descend at random into branches of the vascular network, T cells are introduced uniformly at random across the entire lung surface (Fig.~\ref{fig:systemchart}).  If T cells detect cytokine signal on the local endothelium they remain in tissue and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.  When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.  Unimpeded, infected cells secrete virus for 16.7 hours and then die.

The simulation begins when a single cell in the center of the tissue is infected. After an incubation period, the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type as determined by simulations reported in S2.1 and Table \ref{tab:strains}. Virus diffuses locally, infecting nearby cells, and continuing the cycle. Chemokine diffuses from secreting cells, creating a circle of stimulation around the infected region. After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a steady rate and travel through the vasculature to the tissue.  Further details of the CyCells modeling environment are described in S1.1.

Computational limitations force us to implement our model on the scale of a mouse.  Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).  The surface area of a mouse lung is approximately 100$cm^2$.

In the model (Fig.~\ref{fig:modelchart}), epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.

T cells are described by two states. \texttt{Circulating} T cells emerge from the lymph node at five days post infection and arrive at a random location on the lung's surface after six seconds of travel.  If chemokine is not encountered, the T cell remains a circulating cell, returning to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it converts to \texttt{chemotaxing} and begins following the chemotactic gradient to the infected region. \texttt{Circulating} T cells decay exponentially with an average lifespan of three days.  \texttt{Chemotaxing} T cells move continuously up the gradient through the two dimensional lung endothelium, inducing apoptosis in the \textit{expressing} epithelial cells. \texttt{Chemotaxing} T cells decay exponentially with an average lifespan of two hours.

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \texttt{expressing} epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells at a rate according to the local virus concentration. Chemokine diffuses across the tissue but has no direct effect beyond activating T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of ten after the third day.

Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.  Strain-specific values are shown in Table \ref{tab:strains}.  Further details regarding the model definition are included in S1.2.

@@

Unlike ordinary differential equation (ODE) and DDE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~\ref{fig:variance}).  Therefore, we ran th eCyCells model 50 times for each of the three strains of influenza: avian, seasonal, and 2009 pandemic (Fig.~\ref{fig:variance}).  In every case, the model resulted in viral clearance by day 10 in both the avian and seasonal influenza strains.  Conversely, every simulation of the 2009 pandemic influenza resulted in a runaway infection.

We used $R^2$ values of individual runs versus the average of all runs in order to evaluate the variance of the model.  The $R^2$ ($\pm$ one standard deviation) for avian, seasonal, and pandemic influenza are: 0.957 $\pm$ 0.150, 0.992 $\pm$ 0.013, and 0.998 $\pm$ .002 respectively (i.e. there was little variation between runs).  The three strains show different levels of replication as found in \cite{Mitchell2011}.

Overlaying multiple runs on a single plot reveals a growth rate decline at 4 days post infection, reflecting the addition of IgM.  Subsequently, for each infection the number of infected cells declines quickly after day five post infection due to the CD8 T cell response.  aH5N1 is cleared completely, sH1N1 is more slowly cleared, and pH1N1 clearance at day 6 post infection is halted as replication overtakes clearance.

@@

>>>>

\usepackage{color}
\usepackage[usenames,dvipsnames,table]{xcolor}
\usepackage{ulem}

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin$^{1,\ast}$, 
Stephanie Forrest$^{1}$, 
Soumya Banerjee$^{1}$,
Candice Clay$^{2}$, 
Judy Cannon$^{3}$,
Melanie Moses$^{1}$, 
Frederick Koster$^{1,2}$
\\
\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
\\
\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
\\
\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico, Albuquerque, NM, USA
\\
$\ast$ E-mail: Corresponding drew@cs.unm.edu
\end{flushleft}

@@

Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.  One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.  Here, we focus on T cell navigating from the lymph nodes, through the vascular system to the foci, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.

We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.  The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models. As plaques expand in size, infected regions become more isolated, making it difficult for T cells to cover the infection efficiently.  Further, differences in diffusion rates between chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.  Sensitivity analysis of the model identifies the viral parameters as the most significant factor in determining the course of the resulting infection.  However, the model parameters that collectively determine how efficiently T cells find and clear infected cells are also significant in the model.  These model observations provide explanation of the pandemic H1N1 dynamics that have not been well explained by previous studies.

% To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 H1N1 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.  These data were fit to a differential equation model in order estimate viral and chemokine production rates.

% Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.  This spatially explicit model describes efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.  A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in finding infected cells in large foci.    Thus, we conclude that viral dynamics, rather than chemokine dynamics, dominate the efficiency of T cell recruitment in determining whether the influenza infection is controlled.

@@

Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \cite{Who2009}.  Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood..Pathogenicity is a function of the virus and its interaction with the host immune response.

Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.  However, determining the relative contribution of these myriad experimental components is challenging. We address this challenege with computational modeling, which allows us to study the relative contributions of different aspects of the host immune response, and to account for strain-specific viral dynamics in the model.

Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.  We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.  Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.  Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.  We extend this work to examine how the unique properties of the different strains of influenza may affect this process.

Here, we present a comprehensive agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.  Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models of to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008}.  We use a combination of in vitro generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection to parameterize our model.  Once parameterized with these data, the model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.  These three strains provide a range of documented pathogenicities abd \textit{in vitro} replication efficiencies [TODO citation(s) with pathogenicity data of 3 strains] \cite{Mitchell2011}, which we study in the model.  The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.  Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.

[TODO rewrite to be more specific] The model design allows us to compare between influenza strains how the combination of viral and immune system characteristics affect the strength of the given infection.  The results give us specific insights into how these features [TODO list some] determine the pathogenicity of the different viruses.

The spatial nature of the model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.  Because T cells use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.  The model suggests that the host cells' chemokine response may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.  Finally, the sensitivity analysis isolated the effects of each parameter in the model.  Based on this analysis, each parameter was classified as critical or not.  We find that most parameters can be individually varied across a broad range of values without strong effect on model results.  Further, the parameter classification is consistent across each broad parameter type. Specifically, chemokine-related parameters are stable within biologically plausible ranges.  In contrast, parameters related to viral kinetics have a strong effect on model results.  We conclude that influenza infection dynamics are dominated by the replication properties of the virus, confirming that the higher replication rate of 2009 pH1N1 influenza overwhelmed contributions of the chemotactic signals, leading to uncontrolled \textit{in silico} infection.

% fred used 'obscurred' instead of overwhelmed

% The model predicts the control and clearance of the aH5N1 and sH1N1 strains and predicts a runaway infection for the pH1N1 strain.  These results are consistent with the previously reported replication rates of the three strains of influenza.

%  Infected cells secrete chemokine molecules which then diffuse across the infected region.  This process creates isolated regions, or local maxima, of high chemokine concentrations.

% While the local maxima do move and flatten, their speed is limited by their diffusion and decay rates.

@@

% We are interested in how the interaction between the virus and the host's secondary immune response affects the virulence of the infection.  Because these interactions are difficult and expensive to observe \textit{in vivo}, we create a model \textit{in silico} to represent the system in question.  Because we are interested in individual interactions between the virus and the immune response, as well as the spatial nature of the spreading infection and the subsequent T cell search, we eschew traditional differential equation models \cite{Beauchemin2006a, Bauer2009} in favor of a spatial agent-based model (ABM).

Our modeling methodology follows three steps.  First, we define the system we are going to model.  Second, we build the model with this definition in mind.  Finally, we parameterize the model using empirical data, values from literature, and computational techniques.

%Our model examines the secondary immune response to an influenza infection in a human lung.  Specifically, we are interested in how antigen activated T cells efficiently navigate from the lymph node, through vascular system, and into the lung at the site of infection.

The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).  The lung is represented as a two-dimensional sheet of healthy epithelial cells (S1.3).   Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.  If cytokine signal is detected on the local endothelium the T cell remains in tissue and follows the chemotactic gradient to the FOI.  T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.  Once inside the endothelium, T cells follow the chemokine gradient to localized areas of maximum chemokine concentration.  When a T cell contacts an infected epithelial cell it induces apoptosis (S1.4).  Otherwise, infected cells continue to secrete virus for a fixed time and then die.

The model is initialized with a single infected cell.  After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (S2.1 and Table \ref{tab:strains}).  Virus from expressing cells diffuses locally, infecting neighboring cells. Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI. After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.  These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.

\subsection*{Model Implementation}

The model is implemented using the CyCells software \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).  Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models [TODO add Catherine ref?] \cite{Miller2003, Allan2006, Ingulli2009}.  Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).

Parameters that are independent of the chosen influenza strain are shown in Table \ref{tab:parameters}.  Strain-specific values are shown in Table \ref{tab:strains}.  Further details regarding the model definition are included in S1.2.

[TODO cite Catherine or ourselves] In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.    T cells emerge from the lymph node at five days post infection at a rate of 1,257 cells per hour. T cell travel time from the lymph node to a random location on the lung's surface is six seconds.  If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of three days.  Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells. Chemotaxing T cells decay exponentially with an average lifespan of two hours.

In addition to T cells, the model contains two types of particles: virus and chemokine. Both are produced at constant rates by expressing epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells at a rate proportional to the virus concentration at the location of the cell. Chemokine diffuses across the tissue but has no direct effect beyond recruiting T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day.

\subsection*{Model Parameters}

Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}. These values were taken from the literature when available.  Other parameters were determined experimentally, and some parameters were estimated using biologically plausible ranges.  All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior.  [TODO number of model parameters?]

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended a previously published differential equation model \cite{Mitchell2011} to obtain values for per-cell chemokine production rates (S2.1).  T cell production rates were derived from measured replication rates \textit{in vitro} \cite{Miao2010} using another differential equation model (S2.2). T cell production is assumed to be constant after day 5 [TODO reference or justification for constant rate].

@@

In an effort to obtain per-cell chemokine and cytokine production rates, human epithelial cells were infected with influenza \textit{in vitro}.  Epithelial cell culture and supernatant collection was performed as described in \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table S1).  TNF activity is not incorporated in the model.  Data for other chemokines and cytokines is not shown.

@@

Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).  For the avian and seasonal strains, all model runs cleared the virus by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to runaway infection.

We estimated variance across the runs using the $R^2$ statistic, $\pm$ one standard deviation: 0.957 $\pm$ 0.150 (avian), 0.992 $\pm$ 0.013 (seasonal), and 0.998 $\pm$ .002 (pandemic).  
The $R^2$ was computed with respect to the mean number of infected cells at each simluated time point across all 50 runs. Consistent with earlier results \cite{Mitchell2011}, the time course of the infection varies significantly across the three strains (Fig.~\ref{fig:variance}).

In all studied strains, virus expansion declines at four days post infection (Fig.~\ref{fig:variance}), due to IgM activation.  Subsequently, the simulated CD8 T cell response causes the number of infected cells to decline quickly after day five post infection [TODO p.i. for post infection].  aH5N1 is cleared completely, sH1N1 is cleared more slowly, and at day six post infection, pH1N1 replication dominates the IgM and T cell response.  These results agree with [TODO "previously reported" + citation] \textit{in vivo} differences among the three different strains.

@@

\subsection*{Spatial Effects}
